---
title: "Data Simulation"
author: "Ralph Trane"
date: "Compiled: `r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
bibliography: references.bib
---

## Purpose

As summarized so well by @swanson_partial_2018, by using instrumental variables we can find bounds on the ATE (Average Treatment Effect). Here, we will explore if we can utilize multiple valid IVs to obtain narrower bounds. 

## Simulation: binaries all around

Loosely based on @palmer_nonparametric_2011. The general idea is to simulate data from a number of IVs $Z_1, ..., Z_{n_{iv}}$, a treatment $X$, unmeasured confounder $U$, and an outcome $Y$. In the simplest case, we simulate data as follows:

\begin{align}
Z_i &\sim \text{Bernoulli}(0.5), \\
U &\sim \text{Bernoulli}(0.5), \\
X &\sim \text{Bernoulli}(p_X), \\
Y &\sim \text{Bernoulli}(p_Y), \\
\end{align}

where 

$$
p_X = 0.05 + \sum_{i=1}^{n_{iv}} \alpha_i \cdot Z_i + \gamma_X \cdot U,
$$

and 

$$
p_Y = 0.05 + \beta \cdot X + \gamma_Y \cdot U.
$$ 

### Setup 1: single IV

With a single IV, we look to see how changing coefficients $\alpha_1, \gamma_X, \gamma_Y$ and $\beta$ affect the bounds. 

```{r}
library(bpbounds)
library(glue)
library(dagitty)
library(ggdag)
library(ggrepel)
library(tidyverse)

source("misc.R")

if(!file.exists(paste0(here::here(), "/simulated_data/single_IV_simulation.Rds")) &&
   !interactive()){
  single_IV <- expand_grid(U_on_X = seq(0.01, by = 0.02, length.out = 10),
                           U_on_Y = seq(0.01, by = 0.02, length.out = 10),
                           X_on_Y = seq(0.01, by = 0.02, length.out = 10),
                           indIVs_on_X = seq(0.01, by = 0.02, length.out = 10))
  
  library(furrr)
  plan(multiprocess)
  
  single_IV_simulation <- single_IV %>% 
    mutate(sim_data = future_pmap(list(a = U_on_X, 
                                       b = U_on_Y,
                                       c = X_on_Y,
                                       d = indIVs_on_X),
                                  function(a,b,c,d){
                                    simulate_data(sample_size = 10000,
                                                  U_on_X = a, 
                                                  U_on_Y = b,
                                                  X_on_Y = c, 
                                                  indIVs_on_X = d)
                                  }))
  
  write_rds(x = single_IV_simulation, path = paste0(here::here(), "/simulated_data/single_IV_simulation.Rds"))
  
} 

if(!file.exists(paste0(here::here(), "/simulated_data/res.Rds")) && !interactive()){
  if(!exists("single_IV_simulation"))
    single_IV_simulation <- read_rds(paste0(here::here(), "/simulated_data/single_IV_simulation.Rds"))
  
  library(furrr)
  plan(multiprocess)
  res <- single_IV_simulation %>% 
    mutate(my_bpbounds_res = future_map(sim_data,
                                        ~my_bpbounds(input_data = .x$simulated_data))) %>% 
    mutate(my_bpbounds_res = future_map(my_bpbounds_res, function(x){ 
      x %>% 
        mutate(tidy_output = map(bpbounds, tidy_bpbounds),
               `n Z categories` = map_dbl(bpbounds, function(x) x$nzcats))
    }))
  
  write_rds(res, paste0(here::here(), "/simulated_data/res.Rds"))
} else {
  res <- read_rds(paste0(here::here(), "/simulated_data/res.Rds"))
}


tidy_results <- res %>% 
  select(-sim_data) %>% 
  unnest(my_bpbounds_res) %>% 
  select(-bpbounds) %>% 
  unnest(tidy_output)
```

For fixed $\gamma_X = \gamma_Y = \beta = 0.19$, we vary $\alpha$.

```{r}
tidy_results %>% 
  filter(`Causal parameter` == "ACE", Assumption == "IV Inequality",
         U_on_X > 0.18, U_on_Y > 0.18, X_on_Y > 0.18) %>% 
  ggplot(aes(x = indIVs_on_X, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
  geom_errorbar() + 
  geom_point(color = "red") +
  theme_bw()
```

For fixed $\gamma_X = \gamma_Y = \alpha = 0.19$, we vary $\beta$.

```{r}
tidy_results %>% 
  filter(`Causal parameter` == "ACE", Assumption == "IV Inequality",
         U_on_X > 0.18, U_on_Y > 0.18, indIVs_on_X > 0.18) %>% 
  ggplot(aes(x = X_on_Y, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
  geom_errorbar() + 
  geom_point(color = "red") +
  theme_bw()
```

For fixed $\alpha = \beta = \gamma_Y = 0.19$, we vary $\gamma_X$.

```{r}
tidy_results %>% 
  filter(`Causal parameter` == "ACE", Assumption == "IV Inequality",
         X_on_Y > 0.18, U_on_Y > 0.18, indIVs_on_X > 0.18) %>% 
  ggplot(aes(x = U_on_X, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
  geom_errorbar() + 
  geom_point(color = "red") +
  theme_bw()
```

For fixed $\alpha = \beta = \gamma_X = 0.19$, we vary $\gamma_X$.

```{r}
tidy_results %>% 
  filter(`Causal parameter` == "ACE", Assumption == "IV Inequality",
         X_on_Y > 0.18, U_on_X > 0.18, indIVs_on_X > 0.18) %>% 
  ggplot(aes(x = U_on_Y, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
  geom_errorbar() + 
  geom_point(color = "red") +
  theme_bw()
```

### Setup 2: Multiple independent IVs

```{r}
if(!file.exists(paste0(here::here(), "/simulated_data/multiple_IVs_simulation.Rds")) && !interactive()){
  multiple_IVs <- expand_grid(U_on_X = seq(0.01, by = 0.02, length.out = 10),
                              U_on_Y = seq(0.01, by = 0.02, length.out = 10),
                              X_on_Y = seq(0.01, by = 0.02, length.out = 10))
  
  library(furrr)
  plan(multiprocess)
  
  multiple_IVs_simulation <- multiple_IVs %>% 
    mutate(sim_data = future_pmap(list(a = U_on_X, 
                                       b = U_on_Y,
                                       c = X_on_Y),
                                  function(a,b,c){
                                    simulate_data(sample_size = 10000,
                                                  U_on_X = a, 
                                                  U_on_Y = b,
                                                  X_on_Y = c, 
                                                  indIVs_on_X = seq(0.01, by = 0.02, length.out = 10))
                                  }))
  
  write_rds(x = multiple_IVs_simulation, path = paste0(here::here(), "/simulated_data/multiple_IVs_simulation.Rds"))
  
} 

if(!file.exists(paste0(here::here(), "/simulated_data/res_multiple.Rds")) && !interactive()){
  
  if(!exists("multiple_IVs_simulation"))
    multiple_IVs_simulation <- read_rds(paste0(here::here(), "/simulated_data/multiple_IVs_simulation.Rds"))
  
  library(furrr)
  plan(multiprocess)
  res_multiple <- multiple_IVs_simulation %>% 
    mutate(my_bpbounds_res = future_map(sim_data,
                                        ~my_bpbounds(input_data = .x$simulated_data))) %>% 
    mutate(my_bpbounds_res = future_map(my_bpbounds_res, function(x){ 
      x %>% 
        mutate(tidy_output = map(bpbounds, tidy_bpbounds),
               `n Z categories` = map_dbl(bpbounds, function(x) x$nzcats))
    }))
  
  write_rds(res_multiple, paste0(here::here(), "/simulated_data/res_multiple.Rds"))
} else {
  res_multiple <- read_rds(paste0(here::here(), "/simulated_data/res_multiple.Rds"))
}

Zs_on_X_coefs <- res_multiple %>% 
  select(sim_data) %>% 
  transmute(coefs = map(sim_data, ~.x$coefficients)) %>% 
  unnest(coefs) %>% 
  filter(str_detect(effect, "Z[0-9]+_on_X")) %>% 
  unique() %>% 
  mutate(IV = str_extract(effect, pattern = "Z[0-9]+")) %>% 
  rename(indIVs_on_X = coef)

tidy_results_multiple <- res_multiple %>% 
  select(-sim_data) %>% 
  unnest(my_bpbounds_res) %>% 
  select(-bpbounds) %>% 
  unnest(tidy_output) %>% 
  left_join(Zs_on_X_coefs)
```

For fixed $\gamma_X = \gamma_Y = \beta = 0.19$, we look at how the values for $\alpha_i$ impact the bounds.

```{r}
tidy_results_multiple %>% 
  filter(`Causal parameter` == "ACE", Assumption == "IV Inequality",
         U_on_X > 0.18, U_on_Y > 0.18, X_on_Y > 0.18) %>% 
  ggplot(aes(x = indIVs_on_X, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
  geom_errorbar() + 
  geom_point(color = "red") +
  theme_bw()
```

For fixed $\gamma_X = \gamma_Y = \alpha = 0.19$, we vary $\beta$.

```{r}
tidy_results %>% 
  filter(`Causal parameter` == "ACE", Assumption == "IV Inequality",
         U_on_X > 0.18, U_on_Y > 0.18, indIVs_on_X > 0.18) %>% 
  ggplot(aes(x = X_on_Y, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
  geom_errorbar() + 
  geom_point(color = "red") +
  theme_bw()
```

For fixed $\alpha = \beta = \gamma_Y = 0.19$, we vary $\gamma_X$.

```{r}
tidy_results %>% 
  filter(`Causal parameter` == "ACE", Assumption == "IV Inequality",
         X_on_Y > 0.18, U_on_Y > 0.18, indIVs_on_X > 0.18) %>% 
  ggplot(aes(x = U_on_X, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
  geom_errorbar() + 
  geom_point(color = "red") +
  theme_bw()
```

For fixed $\alpha = \beta = \gamma_X = 0.19$, we vary $\gamma_Y$.


```{r}
tidy_results %>% 
  filter(`Causal parameter` == "ACE", Assumption == "IV Inequality",
         X_on_Y > 0.18, U_on_X > 0.18, indIVs_on_X > 0.18) %>% 
  ggplot(aes(x = U_on_Y, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
  geom_errorbar() + 
  geom_point(color = "red") +
  theme_bw()
```

### Setup 3: multiple independent IVs, and a pair of correlated IVs

Here, we add a pair of correlated IVs, such that we now have

$$
p_X = 0.05 + \sum_{i=1}^{n_{iv}} \alpha_i \cdot Z_i + \alpha_{n_{iv} + 1} Z^1_1 +\alpha_{n_{iv} + 2} Z^1_2 + \gamma_X \cdot U.
$$

$Z^1_i$ is $I(W_i >= 0)$, where 

$$
(W_1, W_2)^T \sim MVN\left((0,0)^T, \begin{pmatrix} 1 & \rho \\ \rho & 1 \end{pmatrix} \right).
$$

```{r}
if(!file.exists(paste0(here::here(), "/simulated_data/multiple_IVs_w_rho_simulation.Rds")) && !interactive()){
  multiple_IVs_w_rho <- expand_grid(U_on_X = seq(0.01, by = 0.02, length.out = 10),
                                    U_on_Y = seq(0.01, by = 0.02, length.out = 10),
                                    X_on_Y = seq(0.01, by = 0.02, length.out = 10),
                                    rho = seq(0.1, 1, by = 0.2))
  
  library(furrr)
  plan(multiprocess)
  
  multiple_IVs_w_rho_simulation <- multiple_IVs_w_rho %>% 
    mutate(sim_data = future_pmap(list(a = U_on_X, 
                                       b = U_on_Y,
                                       c = X_on_Y,
                                       d = rho),
                                  function(a,b,c,d){
                                    simulate_data(sample_size = 10000,
                                                  U_on_X = a, 
                                                  U_on_Y = b,
                                                  X_on_Y = c, 
                                                  indIVs_on_X = seq(0.01, by = 0.02, length.out = 10),
                                                  rho = d, depIVs_on_X = c(0.05, 0.15))
                                  }))
  
  write_rds(x = multiple_IVs_w_rho_simulation, path = paste0(here::here(), "/simulated_data/multiple_IVs_w_rho_simulation.Rds"))
} 

if(!file.exists(paste0(here::here(), "/simulated_data/res_multiple_w_rho.Rds")) && !interactive()){
  if(!exists("multiple_IVs_w_rho"))
    multiple_IVs_w_rho <- read_rds(paste0(here::here(), "/simulated_data/multiple_IVs_w_rho_simulation.Rds"))
  
  
  library(furrr)
  plan(multiprocess)
  
  res_multiple_w_rho <- multiple_IVs_w_rho_simulation %>% 
    mutate(my_bpbounds_res = future_map(sim_data,
                                        ~my_bpbounds(input_data = .x$simulated_data))) %>% 
    mutate(my_bpbounds_res = future_map(my_bpbounds_res, function(x){ 
      x %>% 
        mutate(tidy_output = map(bpbounds, tidy_bpbounds),
               `n Z categories` = map_dbl(bpbounds, function(x) x$nzcats))
    }))
  
  write_rds(res_multiple_w_rho, paste0(here::here(), "/simulated_data/res_multiple_w_rho.Rds"))
} 

if(!file.exists(paste0(here::here(), "/simulated_data/tidy_results_multiple_w_rho.Rds")) && !interactive()){
  if(!exists(res_multiple_w_rho))
    res_multiple_w_rho <- read_rds(paste0(here::here(), "/simulated_data/res_multiple_w_rho.Rds"))
  
  Zs_on_X_coefs_w_rho <- res_multiple_w_rho %>% 
    select(sim_data) %>% 
    transmute(coefs = map(sim_data, ~.x$coefficients)) %>% 
    unnest(coefs) %>% 
    filter(str_detect(effect, "Z[0-9]+_on_X")) %>% 
    unique() %>% 
    mutate(IV = str_extract(effect, pattern = "Z[0-9]+")) %>% 
    rename(indIVs_on_X = coef)
  
  tidy_results_multiple_w_rho <- res_multiple_w_rho %>% 
    select(-sim_data) %>% 
    unnest(my_bpbounds_res) %>% 
    select(-bpbounds) %>% 
    unnest(tidy_output) %>% 
    left_join(Zs_on_X_coefs_w_rho)
  
  write_rds(tidy_results_multiple_w_rho, 
            paste0(here::here(), "/simulated_data/tidy_results_multiple_w_rho.Rds"))
} else {
  tidy_results_multiple_w_rho <- read_rds(paste0(here::here(), "/simulated_data/tidy_results_multiple_w_rho.Rds"))
}
```


```{r}
plot_w_rho <- tidy_results_multiple_w_rho %>% 
  filter(U_on_X > 0.18, U_on_Y > 0.18, X_on_Y > 0.18,
         `Causal parameter` == "ACE", Assumption == "IV Inequality") %>% 
  mutate(IV = factor(IV,
                     levels = paste0("Z", 1:length(unique(IV))))) %>% 
  ggplot(aes(x = IV, y = indIVs_on_X,
             ymin = `Lower bound`, ymax = `Upper bound`)) +
    facet_wrap(~rho, ncol = 2,
               labeller = label_both) + 
    geom_errorbar() + 
    geom_point(color = "red") +
    theme_bw()

plotly::ggplotly(plot_w_rho)
```

